---
title: "introduction"
maintainer: "J. Jang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(JSDNE)
```

###Abstract

Age estimation plays a significant role in forensic anthropology and
bioarchaeology. However, widely-used traditional methods involving
macroscopic observation suffer from subjectivity and statistical bias.
The present research aims to minimize both issues by applying
computational and mathematical approaches. A laser scanner was used to
reconstruct 890 auricular surfaces of adult individuals from three
known-age European skeletal collections. Dirichlet Normal Energy (DNE)
was applied to assess the curvature of the auricular surface and its
relationship with known age-at-death. Ten variables had high
correlations, including total DNE per Total polygon faces, Mean value of
DNE on apex, proportion of polygon faces with DNE of less than 0.0001
and proportion of polygon faces with DNE of over 0.6. The variables were
used to develop age prediction models which are freely available in a
novel R package, JSDNE. The package predicts age mathematically,
objectively, and user-independently. It includes three functions:
principal component quadratic discriminant analysis (PCQDA), principal
component regression analysis (PCR), and principal component logistic
regression analysis (PCLR), which produce age estimates with 91%, 91%,
and 92.9% levels of accuracy, respectively. JSDNE package
(<https://cran.r-project.org/package=JSDNE>) can be downloaded
automatically using install.packages("JSDNE"). The user will obtain the 
estimated results in a few seconds.It is important to note when using this
package that the inputted whole surface and the inputted apex must be written 
in location of x and y, respectively. 

###Introduction

Age estimation, together with sex and stature estimation, is crucial in
constructing the biological profile of human remains. In bioarchaeology,
the biological profile is the foundation for reconstructing past
demographic information, leading to the proper understanding of human
life history. Forensic anthropology uses the biological profile to
identify unknown individuals, helping to investigate cases of missing
persons, mass disasters, genocide, and violent crime (Márquez-Grant,
2015; Moraitis et al., 2014; Rissech et al., 2012). With increased
globalisation, natural disasters, and international terrorism, attention
has been paid to developing more accurate and objective age estimation
methods (Zhang et al., 2022). Age estimation methods have been developed
based on observing bone changes that occur throughout life (Byers,
2007). These bone changes can be divided into developmental and
degenerative changes (Moraitis et al., 2014); developmental changes are
based on the maturation rate of bones, such as epiphyseal ossification
(Schmeling et al., 2004), and tooth eruption (Kumar and Sridhar, 1990).
These are used to estimate subadult age. As for adults who have
completed growth, age is estimated by examining the degree of
degenerative change (Schwartz, 1995), including the pubic symphysis
(Brooks and Suchey, 1990), the sternal ends of the ribs (Işcan et al.,
1984), and the auricular surface (Lovejoy et al., 1985). Among these,
the auricular surface is considered one of the most significant
indicators as it is dense, highly durable and often well-preserved
against decomposition (Rivera-Sandoval et al., 2018). Moreover, it is
argued that degenerative changes on the auricular surface continue after
60 years of age, making it possible to estimate older adults' age more
accurately than other age indicators (Rissech et al., 2012). Even though
age estimation methods have been developed and revised over several
decades, there is still no way to precisely interpret the speed and
degree of change for different individuals. This is because bone changes
are affected by intrinsic and extrinsic factors, such as genes, diet,
socioeconomics, and activity throughout the life, with rates of change
varying between individuals. Therefore, estimating adult age can be
complicated and inaccurate (Miranker, 2016). Age estimation methods
traditionally suffer from two common issues: subjectivity and
statistical bias. Most age estimation methods are based on visual
assessment, which increases subjectivity. When assessing degenerative
changes observers rely on their own expertise and experience to
interpret the morphological features. Thus, the interpretation can
differ depending on the observer, and the estimated result also varies.
Even if the same observer assesses the same bone on different occasions,
the results can be different. These are called interobserver and
intraobserver errors, respectively (Garvin and Passalqua, 2011; Moraitis
et al., 2014). Statistical bias occurs due to differences in the age
structure of the reference and target samples. Fundamentally, the
estimated results for a target sample can be biased by the spread of
ages of individuals that made up the reference sample. This is known as
age structure mimicry (Mensforth, 1990:90). Regional and diachronic
variations can also influence the degree of degeneration displayed by
different morphological features, biasing results towards the reference
sample and causing inaccurate and unreliable results. Overall, these
issues impact the accuracy, reliability, reproducibility and
repeatability of results. To address these limitations, more objective
and less observer-dependent age estimation methods have been sought
(Botha et al., 2019; Curate et al., 2022). Recently, virtual
anthropology has gained attention due to its potential to improve and
revise traditional approaches through the application of computational
imaging technology, such as computed tomography and laser scanning
(Warrier et al., 2023). This allows a mathematical approach to extract
subtle surface changes and more complex visual information not limited
by human vision (Zhang et al., 2022; Warrier et al., 2023). It is also
more user-independent, so the extracted results are objective,
reproducible, and repeatable (Zhang et al., 2022). Currently, the most
widely used mathematical approach in age estimation is curvature
analysis of the joint surface. Curvature analysis is used to evaluate
the degree to which surfaces are undulated or flat, and can therefore be
used to analyse surface texture irregularities (Villa et al., 2015). To
date, most research has focused on the pubic symphysis (Biwasaka et al.,
2013; Stoyanova et al., 2015; Stoyanova et al., 2017; Slice and
Agee-Hewitt, 2015), with only three papers exploring curvature analysis
on the auricular surface (Jang et al., 2024; Štepanovský et al., 2023;
Villa et al., 2015). Villa et al. (2015) used Gaussian curvature
analysis to examine the correlations between five curvature variables
and age: mean, highest, lowest, convex, and flat curvatures. The
variables had moderate correlations with age, though it was noted that
both young and older adults displayed high curvature values; young
individuals because of transverse organisation, and older individuals
due to increased macroporosity. Middle-aged individuals, who generally
showed no transverse organisation or macroporosity (reflecting a flatter
auricular surface), had lower curvature values. However, due to the
sensitivity limitations of the equipment, detailed surface features such
as surface texture and microporosity could not be analysed (Villa et
al., 2015). While Villa et al. (2015) were able to show a correlation
between curvature values and adult age, Štepanovský et al. (2023)
developed this method further to produce age estimation software,
CoxAGE3D, which uses an automatic data mining model. CoxAGE3D is fully
automated, with age estimations produced after the user uploads a 3D
laser scan of an auricular surface to the program. Calculations are
based on multi-population auricular surfaces from five known-age Asian
and European collections, reducing statistical bias levels. While this
method appears to be able to identify a number of age-related
coefficients, accuracy rates were not provided and it is yet to be
subjected to independent validation. Jang et al. (2024) examined
age-related change to the auricular surface using a different type of
curvature analysis, Dirichlet Normal Energy (DNE). This requires a
surface to be converted into polygon faces before DNE is used to measure
how much the surface bends, which it does by quantifying the deviation
of surface energy from a stable state, which is 0 or flat. Each polygon
face (ρ) has two principal curvatures: the largest curvature (v) and the
smallest curvature (υ), and the points at which these meet are called
vertices. Each vertex has a normal, which is the
average value of the adjacent polygonal normal vectors. A
polygon is formed by translating the normal vectors to a common vertex.
This polygon has edge vectors (nν and nυ), which are derivatives of v
and υ (Bunn et al., 2011; Pampush et al., 2016; Winchester, 2016). The
Dirichlet Energy Density (DED) is obtained from the sum of square roots
of edge vectors. Then, DNE is calculated as the total sum of the values
multiplied by DNE and area (Pampush et al., 2016:398):

$$DED(\rho)=(n_v)^2(n_u)^2$$ $$DNE=\sum DED(\rho) \times area(\rho)$$

This makes it possible to identify subtle changes in surface curvature
more sensitively than other methods of curvature analysis (Bunn et al.,
2011; Godfrey et al., 2012). As Jang et al. (2024) tested this method on
archaeological skeletons with no biological information, age was
estimated using the Lovejoy (1985) and Buckberry-Chamberlain (2002)
methods. It was found that four DNE variables correlated with the
estimated age phases, those relating to surface undulation, apical
activity, macroporosity, and the total DNE per total number of polygon
faces. In terms of total DNE per total number of polygon faces, younger
adults had higher values than middle-aged adults due to the presence of
transverse organisation (surface undulation), which occurred more
frequently in the former. Younger adults had the highest DNE values for
surface undulation, with these values falling with increasing age.
Middle-aged adults had lower total DNE values because, as initially
noted by Villa et al. (2015), their auricular surfaces were flatter due
to a lack of transverse organisation or macroporosity. Older adults had
the highest total DNE values and the highest DNE values for
macroporosity and apical activity, which both showed a positive
correlation with age phase (Jang et al., 2024). Jang et al. (2024)
tested the levels of intraobserver error associated with this method and
found very high levels of agreement. However, there was still a degree
of user dependency as the user must manually select a region of interest
(ROI) before DNE values can be calculated (Jang et al., 2024). Moreover,
as this study used archaeological skeletons, the age of the individuals
was not documented, potentially obscuring relationships between DNE
values and age at death. To address these limitations, the current paper
builds on the work of Jang et al. (2024) by reducing the need to select
a ROI manually, and examining the relationship between DNE values and
age-at-death in known-age individuals. Age-related variables will be
identified via three statistical analysis methods: principal component
discriminant analysis (PCQDA), principal component regression analysis
(PCR), and principal component logistic regression analysis (PCLR). The
results have been used to develop an R package, JSDNE, which can be used
to estimate age-at-death in unknown individuals.

###Datasets in the JSDNE package

JSDNE package includes seven datasets: RawData, PCQDA_Train, PCQDA_Test, PCR_Train, PCR_Test, PCLR_Train, PCLR_Test.

####'RawData'
RawData offers the raw information of DNE variables and biological information of auricular surfaces (age and sex). It consists of Age, Cluster1, Cluster2, Collection, Sex, MeanDNE.Apex, MedianDNE.Apex, IQRDNE.Apex, TotalDNE.TotalPolygonFaces, MedianDNE.Whole, IQRDNE.Whole, MeanDNE.Convex, MeanDNE.Concave, Proportion.DNEunder0.0001 and Proportion.DNEover0.6. Cluster1 and Cluster2 is the clustering information for  PCLR and PCQDA, respectively. The number of rows is 890.

####'PCQDA_Train'
PCQDA_Train is the train set of the PCQDA model. It consists of Cluster2, Age, MeanDNE.Apex, TotalDNE.TotalPolygonFaces, Proportion.DNEunder0.0001, and Proportion.DNEover0.6. The number of rows is 704.
This dataset is also used to calculate the Accuracy (ACC), Bias, Average Absolute Deviation (AAD) with PCQDA method. The followed code is used to create PCQDA_Train and PCQDA_Test.

```
PCQDA_data<-subset(RawData,select=c(Cluster2,Age, MeanDNE.Apex, TotalDNE.TotalPolygonFaces, Proportion.DNEunder0.0001, Proportion.DNEover0.6))
sample <- sample(c(TRUE, FALSE), nrow(PCQDA_data), replace=TRUE, prob=c(0.8,0.2))
PCQDA_Train <- PCQDA_data[sample, ]
PCQDA_Test <- PCQDA_data[!sample, ]
```

####'PCQDA_Test'
PCQDA_Test is the test set of the PCQDA model. It consists of Cluster2, Age, MeanDNE.Apex, TotalDNE.TotalPolygonFaces, Proportion.DNEunder0.0001, and Proportion.DNEover0.6. The number of rows is 186.
This dataset is also used to calculate the Accuracy (ACC), Bias, Average Absolute Deviation (AAD) with PCQDA method. 

####'PCR_Train'
PCR_Train is the train set of the PCR model. It consists of Age, MeanDNE.Apex, IQRDNE.Apex, TotalDNE.TotalPolygonFaces, MeanDNE.Convex and Proportion.DNEunder0.0001. The number of rows is 702.
This dataset is also used to calculate the Accuracy (ACC), Bias, Average Absolute Deviation (AAD) with PCR method. The followed code is used to create PCR_Train and PCR_Test.

```
PCR_data<-subset(RawData,select=c(Age, MeanDNE.Apex, IQRDNE.Apex, TotalDNE.TotalPolygonFaces, MeanDNE.Convex, Proportion.DNEunder0.0001))
sample <- sample(c(TRUE, FALSE), nrow(PCR_data), replace=TRUE, prob=c(0.8,0.2))
PCR_Train <- PCR_data[sample, ]
PCR_Test <- PCR_data[!sample, ]
```

####'PCR_Test'
PCR_Test is the test set of the PCR model. It consists of Age, MeanDNE.Apex, IQRDNE.Apex, TotalDNE.TotalPolygonFaces, MeanDNE.Convex and Proportion.DNEunder0.0001. The number of rows is 188.
This dataset is also used to calculate the Accuracy (ACC), Bias, Average Absolute Deviation (AAD) with PCR method.

####'PCLR_Train'
PCLR_Train is the train set of the PCR model. It consists of Age, Cluster1, MeanDNE.Apex, TotalDNE.TotalPolygonFaces, MedianDNE.Whole, IQRDNE.Whole and MeanDNE.Convex. The number of rows is 699.
This dataset is also used to calculate the Accuracy (ACC), Bias, Average Absolute Deviation (AAD) with PCLR method. The followed code is used to create PCR_Train and PCR_Test.

```
PCLR_data<-subset(RawData,select=c(Age, Cluster1, MeanDNE.Apex, TotalDNE.TotalPolygonFaces, MedianDNE.Whole, IQRDNE.Whole, MeanDNE.Convex))
sample <- sample(c(TRUE, FALSE), nrow(PCLR_data), replace=TRUE, prob=c(0.8,0.2))
PCLR_Train <- PCLR_data[sample, ]
PCLR_Test <- PCLR_data[!sample, ]
```

####'PCLR_Test'
PCLR_Test is the test set of the PCR model. It consists of Age, Cluster1, MeanDNE.Apex, TotalDNE.TotalPolygonFaces, MedianDNE.Whole, IQRDNE.Whole and MeanDNE.Convex. The number of rows is 191.
This dataset is also used to calculate the Accuracy (ACC), Bias, Average Absolute Deviation (AAD) with PCLR method.

###Materials

Data for the current study was collected from three European anatomical
skeletal collections; the Luís Lopes Skeletal Collection, the 21st
Century Identified Skeletal Collection, and the CAL Milano Cemetery
Skeletal Collection. The Luís Lopes Skeletal Collection is curated by
the National Museum of Natural History and Science (MUHNAC), Lisbon,
Portugal. It contains 1,692 skeletons in total, with biographical
information available for 699 individuals, including age-at-death, place
of birth and residence, and cause of death. The remains were collected
from a modern cemetery in Lisbon between 19th and 20th centuries
(Cardoso, 2006). The 21st Century Identified Skeletal Collection is
curated by the University of Coimbra, Portugal. This contained 159 adult
individuals, ranging in age from 29 to 99 years at death. Most
individuals were Portuguese, dying between 1995 and 2008. Biographical
information, including sex and age-at-death, is available for the entire
collection, with documentation relating to the cause of death available
for 12 individuals (Ferreira et al., 2014). Lastly, the CAL Milano
Cemetery Skeletal Collection is curated by the Laboratorio di
Antropologia e Odontologia Forense (LABANOF) in the Department of
Biomedical Sciences for Health of the University of Milan (Italy) and
part of the Collezione Antropologica LABANOF (CAL), Italy. It contains
2127 skeletons from Cimitero Maggiore, Cimitero di Lambrate, and
Cimitero di Baggio. Biographical information, such as sex, age-at-death,
date of birth and death, cause of death, and pathology (if it is
specified), were all documented. The individuals died between 1910 and
2001, but most of them died after 1980 (Cattaneo et al., 2018). The
current research consists of 890 (402 females, 488 males) adult
individuals with a known age-at-death from these three collections. 
All 890 individuals had fully preserved auricular surfaces that displayed 
no taphonomic damage or pathological
conditions.

###Methods

Preparation and importing of reconstructed models Auricular surfaces
were scanned using an EXScan SP laser scanner (resolution: 1.3
megapixels). Each surface was scanned from a minimum of five different
angles and the resultant images were combined using the software
EXScanS_v3.1.0.1 to produce a single 3D model which was saved as a .ply
file. Each .ply file was then edited manually in Meshmixer to produce
two ROIs; the auricular surface and the apical area. The auricular
surface was isolated using the outer margin as a guide (following the
criterion of Jang et al., 2024). The apical area was delineated by
measuring polygons within a 1cm distance from both the left and right
sides of the terminated point of the arcuate line. The 3D models were
imported into R studio with the vcgPlyRead function from Rvcg package
(Schlager, 2015).

Analysis Principal component analysis (PCA), quadratic discriminant
analysis (PCDQA), and regression analysis (PCLR) were used to identify
DNE variables that were highly correlated with known age-at-death. These
correlations were then used to develop a model that estimates an unknown
auricular surface's age-at-death. In order to make this model available
to other researchers, an R package, JSDNE, was created, which requires R
version 4.3.2 (R Core Team, 2024). JSDNE can be installed automatically
by operating the install function, install.packages("JSDNE"). Once the
package is installed, it can be imported by the library function,
library(JSDNE). Four packages are automatically imported to the user's
library: dplyr (Wickham et al., 2019), MASS (Venables and Ripley, 2002),
molaR (Pampush et al., 2016), and nnet (Venables and Ripley, 2002).
After importing the .ply files with the vcgPlyRead function, the
auricular surface .ply file should be replaced with x and the apical
area with y. If oppositely applied, incorrect results will be
calculated. The JSDNE package consists of three functions to estimate
age. The PCQDA function can assign an individual to one of
four broad age ranges: phase 1 (20-44yrs), phase 2(31-74yrs), phase 3 (63-93yrs), phase 4 (over 76) The PCR function provides an output of
estimated age in years and the standard error (8.8yrs). The PCLR function also
assigns an individual to a broad age range, but with a more limited
range of under 67 years (phase 1) or over 63 years (phase 2).


###Results

Age-related variables Initial analyses identified ten DNE variables that
were highly correlated with age-at-death in the known-age sample (Table
3 and Figure 3). Nine of these variables (total DNE per total number of
polygon faces, median value of DNE on whole surface, interquartile range
(IQR) on whole surface, mean value of DNE on convex and concave
surfaces, proportion of polygon faces with DNE of over 0.6, and mean,
median and IQR of DNE on apex) had high positive correlations, while the
proportion of polygon faces with DNE of less than 0.0001, had a high
negative correlation.

Identifying the suitable variables and the cluster The best combination
of variables was chosen on the basis of three criteria: (1) high total
accuracy, (2) high accuracy in younger auricular surfaces and older
auricular surfaces, and (3) low bias. For PCR analysis, an additional
criterion was used: low standard error. For the PCQDA analysis, four
variables provided the best combination: proportion of polygon faces
with DNE of less than 0.0001, proportion of polygon faces with DNE of
over 0.6, mean value of DNE on apex, and total DNE/total polygon faces.
Grouping estimated ages into four age phases (Table 4 and Figure 4). The
age intervals of each phase are 24 years (phase 1), 43 years (phase 2)
and 30 years (phase 3), with an open-ended phase 4. For PCR analysis,
seven variables provided the best combination: total DNE per total
number of polygon faces, mean value of DNE on apex, IQR value of DNE on
apex, mean value of DNE on convex surface, proportion of polygon faces
with DNE of less than 0.0001. As this is a regression model no age phase
was produced. Instead, the output provides an estimated age in years
with a standard error of ±10.1 yrs. Finally, the PCLR analysis was
developed using five variables: total DNE per total number of polygon
faces, median value of DNE on whole surface, IQR of DNE on whole
surface, mean value of DNE on convex surface, and mean value of DNE on
apex. Estimated age was divided into two phases to determine if the
auricular surface is under 67 years or over 63 years (Table 5 and Figure
5). The age phases for each of the three types of analysis were
designated in order of the lowest minimum value of the age range. The
minimum values of age phases 1,2,3 and 4 in PCQDA were 20,31,63 and 76,
respectively, and minimum values of age phases 1 and 2 in PCLR were 20
and 63, respectively. When converting the variables to principal
components, the first and second principal components contained 98%,
96%, and 96% information in PCQDA, PCLR, and PCR methods, respectively.
Thus, the first and second principal components were used as variables.
Then, the statistical models were developed using cross-validation. The
training and test sets contained 80% and 20% of the data set randomly.


###Accuracy, bias, and the average absolute deviation

PCQDA The training set, test set, and combined set showed similar
results. Accordingly, the results were described based on the combined
set. Accuracy was defined as JSDNE age estimates that fell within the
age phase that contained the known age-at-death. Bias was calculated as
the mean value of the subtracted values between the mean age of the
estimated age phase and the known age at death. The average absolute
deviation (AAD) was the mean value of the absolute values subtracted
between the mean age of the estimated age phase and the known age at
death:

$$Bias=\sum(mean \ age \ of \ the \ estimated \ age \ phase - known \ age \ at \ death)/N$$
$$AAD=\sum|mean \ age \ of \ the \ estimated \ age \ phase - known \ age \ at \ death|/N$$

###Summary of the Results of PCQDA model

The total accuracy and the average absolute deviation (AAD) are 91% and
9.58 years, respectively. This demonstrates that the PCQDA
function estimates age with a high level of accuracy and precision.
Phase 2, which estimated individuals to be over 31 and under 74 years of
age, had the lowest accuracy (88.8%) and the highest AAD (11.52 years).
The highest accuracy and the lowest AAD were shown in phase 1 (20-44
years of age), with 97.8% and 5.53 years, respectively. While the
cluster analysis identified this four-phase grouping system as the best
fit for the prediction model, levels of accuracy were also examined
using 10-year age intervals, in order to compare the relative
applicability of each of the functions across age cohorts. This showed
that there was a tendency to underestimate the age of older individuals
(those in their 60s and older) and overestimate the age of younger and
middle aged individuals (in their 50s and younger). Individuals in their
20s were estimated with the lowest level of accuracy (65.9%). The
highest AAD (94.9 years) was in their 30s. The highest accuracy levels
were seen for individuals estimated to be in their 40s (98.2%).

###Summary of the Results of PCR model

PCR Accuracy was defined as JSDNE age estimates that fell within the
standard error (8.8yrs) of the known age-at-death. The total accuracy
and ADD are 91% and 6.46 years, respectively. When collated
into 10-year age intervals, the accuracy levels range from poor to
moderate, but this is due to the high precision of the age estimate in
the majority of intervals. Individuals in their 20s and over 80 were
predicted with the lowest levels of accuracy (52.4% and 56.3%), and
highest levels of AAD (9.37 years and 9.4 years). The highest accuracy
(92.2%) was found for individuals estimated to be in their 50s. As with
the PCQDA function, there was a tendency for younger and middle aged
adults (those in their 40s and younger) to be overestimated, while the
age of individuals over 50 was underestimated.

###Summary of the Results of PCLR model

PCLR Accuracy was defined as JSDNE age estimates that fell within the
age phase that contained the known age-at-death. PCLR produced a high
level of accuracy overall (92.9%), and high precision (9.93 years). Phase 1 has higher accuracy (95.2%) than phase 2 (90.2%). However, phase 2 has higher precision than phase 1, indicated by the
lower AAD value (11.48 years vs. 8.09 years). When classifying
individuals with 10-year age intervals, the highest accuracy
was shown for individuals in their 30s (100%), and the lowest accuracy
for individuals in their 60s (80.6%). In terms of precision, individuals
in 20s had the lowest precision (ADD: 20.45 years), and those in their
40s had the highest precision (ADD: 2.68 years). Yet again, individuals
under 40 years of age tended to be overestimated, while individuals over
40 were underestimated.

###Discussion

Initial analyses involving the DNE values and known age-at-death
revealed ten DNE variables that were highly correlated with age (Table 3
and Figure 3). The majority of these variables had positive
correlations, demonstrating that the auricular surface becomes more
curved with increasing adult age. The only variable to show a negative
correlation was the proportion of polygon faces with DNE of less than
0.0001, which demonstrates that flatter areas of the auricular surface
(the number of polygons that are mathematically flat as a product of nv
and nu), become reduced with increasing age as the surface becomes more
irregular. These findings are in agreement with traditional methods of
age estimation, which have long been aware that areas of macroporosity
and osteophytic growth become more visible on the auricular surfaces of
older individuals (Buckberry and Chamberlain, 2002; Lovejoy et al.,
1985). Previous research with curvature analysis has suggested that
young adults have high curvature values (Jang et al., 2024; Villa et
al., 2015), a finding which is not echoed here. It is well known that
the auricular surfaces of younger adults display distinctive areas of
transverse organization over a large proportion of the surface
(Buckberry and Chamberlain, 2002; Lovejoy et al., 1985) and it was
previously thought that this feature contributed to the high levels of
curvature reported by Jang et al. (2024) and Villa et al. (2015).
However, while transverse organization is visible on the reconstructed
models used in the current study, it appears to have had no measurable
effect on curvature. It is possible that results for younger adults in
Jang et al. (2024) and Villa et al. (2015) are highly affected by the
presence of subchondral defects (which produce very high DNE values
(\>0.6)). While subchondral defects are present in all age groups, they
have a limited on impact on the variance and median DNE values when the
sample is large. However, when the sample is small, the inclusion of
several individuals with subchondral defects can increase the variance
dramatically and raise median DNE values for the entire group, leading
to improper generalisation due to the small sample size. These results
demonstrate that variables relating to a number of the morphological,
degenerative features used in traditional age estimation methods can be
identified through DNE analysis. However, instead of relying on the
experience of the observer, DNE can quantify mathematical aspects of
surface geometry that are correlated with age, reducing the subjectivity
and user-dependence. The JSDNE R package uses DNE variables that are
correlated with age-at-death to estimate age for unknown auricular
surfaces. Three functions are available; PCQDA, PCR, and PCLR. The total
accuracy of the PCQDA function produces the highest level of accuracy
(91%), likely a result of wide age ranges of each age phase (Millán et
al., 2013). The PCLR function also produces a high level of accuracy
(92.9%), but is used to determine whether an individual is under 67 or
over 63 years of age. The PCR function produces a narrower age range
(SE: 8.8yrs) with the high level of accuracy (91%). These results are
similar to the accuracy rates gained by the majority of researchers who
carry out validation studies of the Buckberry and Chamberlain (94% -
86%) and Lovejoy (63% - 27%) methods of traditional analysis (Hughes et
al., 2024; Milán et al., 2013; Rissech, et al., 2012). However, while
the traditional methods rely on subjectivity and the researchers own
experience in assigning an individual to an age phase (Štepanovský et
al., 2023), the JSDNE results are completely objective and have
excellent levels of repeatability. When choosing which function to use,
the nature of the research must be considered. PCQDA and PCLR may be
more suitable for estimating age in archaeological individuals as these
functions can identify broad age ranges that correspond to various
stages of the lifecourse with moderate to high levels of accuracy. For
forensic cases, which require a more precise estimate of age-at-death,
the PCR function may be more appropriate. It is also possible to use all
three functions in conjunction with one another. Although able to
provide an objective and user-independent estimate of age-at-death,
these results still demonstrate a degree of bias, as the majority of the
functions tended to overestimate the age of younger individuals and
underestimate the age of older individuals. This is a common tendency
that has been observed in other studies (Falys et al., 2006; Moraitis et
al., 2014; Navitainuck et al., 2022; Rissech et al., 2012;
Rivera-Sandoval et al., 2018; Štepanovský et al., 2023), and relates to
the age structure of skeletal samples used to develop the methods.
Generally, the number of younger and older adults is fewer than those
who are middle-aged, meaning that a wider range of variation is captured
for the middle-aged auricular surfaces and the resulting age estimates
are also biased towards this age group. There is currently no way to
alleviate this problem with the known-age skeletal collections that are
available worldwide.

###Conclusion

The present research successfully identified ten geometric variables of
the auricular surface that were associated with known age-at-death and
used these to create an objective, mathematical age estimation method
that is available as a novel R package, JSDNE
(<https://cran.r-project.org/package=JSDNE>). This estimates age using
three functions which offers the high levels of accuracy (91-92.9%).
This method does not require a great degree of experience to gather and
interpret the results, only familiarity with R studio and access to a
laser scanner. Moreover, the results can be applied to research in
bioarchaeology and forensic anthropology.

###Acknowledgements

We would like to appreciate to Dr. Maria Susana De Jesus Garcia with the
Luís Lopes Skeletal Collection, curated by the National Museum of
Natural History and Science (MUHNAC), in Lisbon, Portugal, Dr. Maria
Teresa Ferreira with the 21st Century Identified Skeletal Collection,
curated at the University of Coimbra, in Portugal, Dr. Mirko Mattia and
Dr. Cristina Cattaneo with the CAL Milano Cemetery Skeletal Collection,
curated at the University of Milan, in Italy, allowing to access the
collections.


###Code Summary
#### `PCQDA_result()`

PCQDA method involves four age phases: age phase 1 (age range:20-44), age phase 2 (age range:31-74), age phase 3 (age range:63-93), and age phase (age range: over 77). The result is the estimated age phase and its age range. This is the example how to operate the PCQDA model. The WholeSurface and Apex files are imported ply file of the scanned auricular surface. Both file can be downloaded though https://github.com/jisunjang19/cran-JSDNE (in data folder). DNE variables (MeanDNE.Apex, TotalDNE.TotalPolygonFaces, Proportion.DNEunder0.0001, Proportion.DNEover0.6) are calcaulted automatically in the PCQDA_result() function. Then, the function is estimated the age phase.The output is the estimated age phase and age range.

```{r}
library(JSDNE)
PCQDA_output <- PCQDA_result(WholeSurface, Apex)
```

#### `PCR_result()`
PCR method estimates the age as the regression analysis. The result is the estimated age and the standard error (8.8 yrs).

This is the example how to operate the PCR model. The WholeSurface and Apex files are imported ply file of the scanned auricular surface. Both file can be downloaded though https://github.com/jisunjang19/cran-JSDNE
(in data folder). DNE variables (MeanDNE.Apex, IQRDNE.Apex, TotalDNE.TotalPolygonFaces, MeanDNE.Convex, Proportion.DNEunder0.0001) are calcaulted automatically in the PCR_result() function. Then, the function is estimated the age. The output is the estimated age with standard error (8.8 yrs).

```{r}
library(JSDNE)
PCR_output <- PCR_result(WholeSurface, Apex)
```

#### `PCLR_result()`
PCLR method involves two age phases: age phase 1 (age range: under 67), and age phase 2 (age range: over 63). The result is the estimated age phase and its age range.
This is the example how to operate the PCLR model. The WholeSurface and Apex files are imported ply file of the scanned auricular surface. Both file can be downloaded though https://github.com/jisunjang19/cran-JSDNE (in data folder). DNE variables (TotalDNE.TotalPolygonFaces, MeanDNE.Apex, MeanDNE.Convex, MedianDNE.Whole, IQRDNE.Whole) are calculated automatically in the PCR_result() function. Then, the function is estimated the age phase. The output is the estimated age phase and age range.

```{r}
library(JSDNE)
PCR_output <- PCR_result(WholeSurface, Apex)
```
