---
title: "introduction"
maintainer: "J. Jang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(JSDNE)
```

## Introduction
JSDNE package has been developed as a novel adult age estimation method using computational and mathematical approaches. It uses the 3D reconstructed whole auricular surface and apex, saved as a .ply file. It is recommended that those ply files be inputted using the Rcvg package. Once the files are inputted, the curvatures are calculated by Dirichlet Normal Energy (DNE), and the DNE-curvature variables are created in the code. The package provides three age estimation methods: PCQDA, PCR, PCLR. The user will obtain the estimated results in a few seconds.It is important to note when using this package that the inputted whole surface and the inputted apex must be written in location of x and y, respectively.


### `PCQDA_result()`
PCQDA method involves four age phases: age phase 1 (age range:20-54), age phase 2 (age range:31-74), age phase 3 (age range:63-93), and age phase (age range: over 77). The result is the estimated age phase and its age range.

### `PCR_result()`
PCR method estimates the age as the regression analysis. The result is the estimated age and the standard error (12.4 yrs).

### `PCLR_result()`
PCLR method involves two age phases: age phase 1 (age range: under 67), and age phase 2 (age range: over 63). The result is the estimated age phase and its age range.
